Gradual Soundness: Lessons from Static Python

### 2022-04-26 pj-7.1/response.txt

Thank you all for the detailed reviews. We have made the major changes that
we promised in our edit plan and followed many of your other suggestions.

For reference, here are the major changes:

 1. (Section 4) Clearly explain the purpose of the paper's model and the Redex
    model, and explain how they relate to one another and to Static Python. In
    particular, address the following specific points from the reviews:
    - (#45A) Explain how Redex's random testing helps to check type soundness.
    - (#45B) Explain "soundness" and "property-based soundness tests" more
      clearly.
    - (#45B, #45C) Clearly say that the model's typing is not algorithmic and
      that Redex's typing is algorithmic. Explain how and why the model is
      different.
    - (#45A, #45B) Explain the semantics of objects and method calls in more
      detail.

  2. Explain gradual soundness in more detail throughout the paper.

  3. (Section 7) Extend the related work and be sure to compare this work to
     prior work.

Below, please find point-by-point replies with page and section numbers to
show where the revised paper has changed.

> Review #45A
> ===========================================================================
>
> [... clipped ...]
>
> ## comments
>
> I like the paper, it is generally well readable. There are a few places,
> where I'd expect more guidance from the authors, rather than expecting the
> reader to interpolate.
>
> ### Table 1
>
> The caption is misleading. The data is the cost of creating and using
> shallow/concrete dictionary types in typed and untyped contexts.  I was
> initially confused by the 0 entries. I'd prefer you put --- to indicate that
> no casts is needed.  You should indicate that $N$ is the length of the
> dictionary.

@@

> ### p10, 2
>
> *only if*
> Is that an `iff`?

No, the "only if" is intentional. An "iff" would be correct, but the paragraph
in which this sentence appears is about one direction in particular.

> ### p10, -17
>
> *Redex's random testing ... to check type soundness*
>
> That seems a bit shaky. What's the baseline / ground truth / oracle for these
> tests?
> In the end, all that testing only gives you a probabilistic guarantee.
> Can you give an estimate of this probability? (I mean, quantitative)

These are property-based tests. Our model defines a soundness property for terms
and uses Redex to generate random terms in the hope of finding a counterexample.
So, the soundness property is the ground truth; these random tests are all
checking soundness. The basic outline is: if a random term is well-typed and
terminates within a finite number of steps, then it must either reduce to a
well-typed value or an error. (To get thousands of well-typed and terminating
terms, we have Redex generate tens of thousands of random terms.)

@@

One could imagine a probabilistic guarantee about what kinds of terms the tests
explore, but the guarantee would depend on two complicated parameters: our
model's grammar and Redex's algorithm for exploring grammars (Klein & Findler,
SFP 2009). These parameters make it difficult to formulate a guarantee. On top
of that, **explaining** the guarantee to readers would be a huge challenge.

In the end, although property-based testing is not as convincing as formal
verification, it is still a helpful tool for finding bugs. Indeed, we found
several bugs in our model thanks to these property-based tests (none of which
had implications for Static Python itself).

@@

Method calls in Redex are standard; they work like calls in (Static) Python.
The confusion here is due to our over-simplified model in the submission, which
gave the impression that only a class can receive a method call.

> ### p24
>
> references [23] and [24] and [41] have broken DOIs

Fixed.

> * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * *
>
>
> Review #45B
> ===========================================================================
>
> [... clipped ...]
>
> Detailed comments:
>
> Page 3 Mechanization: you should explain better what do you mean by
> "soundness". I thought of standard soundness in the sense that "well-typed
> programs do not go stuck", but here you say "property-based soundness tests",
> what do you mean more precisely?

@@

We disagree. The section is a proof sketch that ill-typed values cannot "sneak
in" through the dynamic type.

> Section 6.3 I miss a conclusion/summary of the analysis here.

We've added a conclusion section.

> * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * *
>
>
> Review #45C
> ===========================================================================
>
> [... clipped ...]
>
> Assessment and comments (for authors)
> -------------------------------------
> The language design mixing concrete and shallow types is interesting.
> It seems that the language has been accepted by engineers and used for
> a large software project.  Nevertheless I have a few reservations to
> accept the paper in its current form.
>
> The paper puts emphasis on "gradual soundness" at the beginning but,
> after reading the paper, I failed to see what it really was.  The
> phrase is never mentioned on page 3 and afterwards.  This is
> unfortunate because this is the phrase used even in the title!

### 2022-04-26 pj-7.1/paper.tex
pj: edit pass paper + response

%% TODO 2022-01-25 recount tests
\newcommand{\numbenchmark}{3}
\newcommand{\numSPtest}{265}
\newcommand{\numSPtotal}{802} %% SP commit 6d61575 from Jan 9
\newcommand{\numSPdiff}{537} %% (\numSPtotal - \numSPtest)
\newcommand{\CPUchange}{3.7\%}

@@

This paper reports an exception to the rule among industry-made gradual type systems.
The \SP{} team at Instagram has developed a sound type system for a subset of
Python along with a runtime system that uses soundness to drive optimizations.
The language is staged to let programmers choose between easy migrations and
full-strength optimizations, a design that we call \emph{gradual soundness}.
To a first approximation, there are three main stages:
\begin{enumerate}
  \item
    A full language of \emph{shallow types\/} that describe idiomatic
    Python code at a coarse granularity.
    Adapting an untyped module to use shallow types requires at most a few local code
    changes.
  \item
    A second level of \emph{concrete types\/} describes generic data structures
    that check the types of their elements.
    If programmers modify their code to build concrete structures instead of Python
    ones (a potentially nonlocal change), then \SP{} can perform deeper optimizations.
  \item
    A third set of progressive types enable further optimizations.
    In particular, \emph{primitive types\/} allow for unboxed arithmetic,
    which has been critical in practice.
\end{enumerate}

@@

  \item
    \emph{Mechanization}.
    To validate the soundness of \SP{}, we mechanized a core language in PLT Redex
    and ran both property-based soundness tests and over 200 tests adapted from the \SP{} regression suite.
    The modeling effort revealed 20 significant issues in \SP{},
    four of which were soundness bugs.

@@

This paper begins with an informal description of \SP{} in two parts.
First, we present a user-oriented summary of the language~(\cref{s:tour}).
Second, we present the key ingredients of the runtime system, Cinder,
that supports \SP{}~(\cref{s:runtime}).
\Cref{s:model} uses a small formal model to introduce our Redex mechanization
and to convince readers that \SP{} is based on a sound core.
\Cref{s:impl} notes important aspects of \SP{} that are not covered by the model.
\Cref{s:eval} evaluates \SP{}; it reports our experience with the language in
production and on public microbenchmarks.
The paper concludes with related work~(\cref{s:related})
and a final summary~(\cref{s:conclusion}).

@@

\SP{} is part of a large codebase that includes a type system, a tailored
bytecode, and a method-based JIT compiler.
In essence, \SP{} is the type system for an entirely new language.
The interface that it offers to programmers, however, replicates the
standard Python experience.
\SP{} runs Python 3.8 programs with minimal changes to their
behavior~(\cref{s:not-python}),
and it compiles code on-the-fly to support existing
IDEs and developer tools.
The advanced features of the \SP{} type system are offered on an opt-in basis
and arranged so that programmers can begin adding types one module at a time.

@@

The \SP{} type system is a unique synthesis of ideas from the gradual typing literature
and prior work on types for Python.
Its syntax for types is based on the definitions in \PEP{}~\cite{pep484}.
Its static semantics for types is inspired by optional type checkers; in
particular, Pyre~\cite{pyre} and mypy~\cite{mypy}.
And its strategy for run-time checks is adapted from Nom~\cite{mt-oopsla-2017,mt-oopsla-2021},
a research language with compelling performance results.
\SP{}'s novelty comes from the following engineering constraints:
\begin{itemize}
  \item
    Performance is the bottom line.
    At the end of the day, \SP{} needs to make code run faster in production.
  \item
    Critical code often relies on first-order functions and objects,
    at least in the Instagram web server.
    Precise types for higher-order functions, for first-class classes, and even for
    locally-defined functions and classes are thus a low priority; instead, the
    dynamic type serves as a coarse approximation.
  \item
    The codebase in which \SP{} is being applied makes heavy use of
    \PEP{} type annotations and is regularly analyzed by Pyre.
    \SP{} is therefore compatible with the \PEP{} syntax to reduce the
    adoption burden.
    %% depth-first approach? (the alternatives are not very clear ... depth vs breadth?)
  \item
    A module-level granularity is acceptable.
    Once \SP{} is enabled for a module, it compiles all code in that module
    including expressions that have the dynamic type~(\cref{s:type-dynamic}).
    If this behavior is a problem for certain expressions, it is easy enough to move
    them to an untyped Python module.
\end{itemize}
%
In general, the types that \SP{} implements all enable significant optimizations
and can be validated quickly with casts.
Types that do not meet these criteria are deferred to the mature optional
type checkers that already exist for Python.
In fact, \SP{} has no short-term plans to support all \PEP{} types.

@@

\end{enumerate}
%
With this context in mind, \SP{} keeps track of whether each class is typed or untyped.
Each typed class can furthermore assume that if a method has a precise (non-dynamic)
type, then all typed overrides are subtypes of this type.
\SP{} can therefore optimize dispatch from typed callers to typed methods
and use checked dispatch for other combinations.
\SP{} can also optimize reads from typed fields.
One extra step arises when an untyped method overrides a typed method.
In this case, \SP{} creates a wrapper around the overriding method to check
that it computes type-correct results.
% Because these wrappers perform a simple first-order check, they
% are supported by an efficiently implementation that cooperates with
% \SP{}
The wrappers are handled efficiently by a tailored implementation of vtables~(\cref{s:vtable}).

@@

For performance-critical applications, \SP{} includes a set of primitive types
that describe booleans and sized numbers.
Example types include \code{int64}, \code{uint64}, \code{double}.
There are also two special datatypes, \code{Array} and \code{Vector}, that store primitives.
The purpose of these types is to allow unboxed values and arithmetic at
runtime.
Consequently, their static semantics is a \emph{progressive}~\cite{pqk-onward-2012}
refinement over the semantics of basic Python types:
they permit fewer behaviors statically to enable the unboxing.
Operations such as \code{and} (boolean conjunction), for example, do not accept
a mix of primitive and non-primitive arguments.
These primitive types are another main aspect of gradual soundness.

@@

Relative to Python, \SP{} makes a few behavioral changes to enable
sound types and strong optimizations:
class attributes must be set in an \code{\_\_init\_\_} method to
ensure predictable layouts (all \PEP{} checkers agree on this);
typed attributes are resolved statically, and thus cannot be overridden with
descriptors or properties;
module-level variables cannot be mutated;
typed fields 
and
multiple inheritance of either typed classes or untyped children
of typed classes is forbidden.

@@

All CPython 3.8 instructions have the same behavior in Cinder.
Cinder adds instructions to help with one of three tasks:
expressing runtime checks,
initializing concrete-typed values,
or efficiently performing a standard action.
As an example of the third kind, the \code{FAST\_LEN} instruction quickly
computes the length of a built-in value, which helps to optimize for loops.
\SP{} uses type information to decide where this instruction is appropriate
and applies it as an optimization.

@@

The concrete versions of built-in data structures come with both a type and an
implementation.
The implementation provides the same interface as the built-in but uses a
type tag to reject certain operations.

@@

Cinder includes a JIT (just-in-time) compiler for its bytecode.
At the moment, programmers enable the JIT by supplying a list of method names
to the compiler.
Any method can be JIT-compiled whether or not it uses \SP{} types,
even though fewer types usually leads to fewer optimizations.

@@

The modelling process helped uncover several critical issues in \SP{}.
This is a very important payoff given that \SP{} is running in production at
Instagram.
Overall, we made 25 bug reports~(\cref{a:github-issues}) to the \SP{} team via GitHub.
Four of these were soundness issues, one of which we could exploit to raise
a segmentation fault.
All but one of these soundness bugs have been fixed.
Of the remaining issues, five were relatively minor; these dealt with confusing error messages
and incorrect tests.
The remaining 16 issues report bugs in language design and implementation.
The \SP{} team has acknowledged these as bugs by applying a specific GitHub label: \lstinline{sp-correctness}.

%% critical soundness issues:
%% https://github.com/facebookincubator/cinder/issues/55
%% https://github.com/facebookincubator/cinder/issues/53
%% https://github.com/facebookincubator/cinder/issues/39
%% https://github.com/facebookincubator/cinder/issues/36

@@

The rest of this section illustrates our Redex model using a small
formalization.
%% TODO EXAMPLE simplification
\Cref{s:surface} presents a surface syntax to highlight the boundaries between
typed and untyped code.
\Cref{s:eval-types} explains how types get enforced at run-time.
\Cref{s:ts} argues that the overall approach toward boundaries is sound.

@@

\Cref{f:surface-types} presents a core syntax for programs.
A program is a sequence of statements;
a statement defines a variable, function, or class.
These definitions may only appear on the top level
and they all require type annotations.
Functions must have one positional argument.
Classes must declare one parent (either $\sptobject$ or another class), one field, and one method.
Expressions describe values and simple computations.
The basic values are the none value, integers,
booleans (which are the integers \code{0} and \code{1}), strings, and the top object.
There are two data structures: dictionaries and checked dictionaries~(\cref{s:checked-type}).
Other expressions describe function calls, dictionary reads and writes, object field reads and writes,
and method calls.

In Python, the syntax for expressions overlaps with the syntax of types.
For example, the Python name \code{None} is both an expression and a type,
and the \SP{} names \code{chkdict} and \code{CheckedDict} are synonyms.
\Cref{f:surface-types} does not follow these standards; instead it keeps the two syntaxes distinct.
Expressions use lowercase names (\code{none}, \code{chkdict}) and types use capitalized
names (\code{None}, \code{CheckedDict}).

Types $\sptype$ include the dynamic type, types for the basic values, one type $\sptclass$ for every
user-defined class, and union types.
We choose to write the dynamic type as $\sptdyn$ for simplicity; technically,
\SP{} does not have a surface syntax for its dynamic
type and instead allows missing and unknown types~(\cref{s:type-dynamic}).
We assume that all unions are written in a flat and simplified form, e.g., that
$\sptunion{\sptint, \sptunion{\sptdyn, \sptclass_0}}$ would be normalized to $\sptdyn$.

@@

\SP{} evaluation types can all be enforced fully and efficiently
with casts.
\Cref{t:cast} describes the casts in detail.
Most call for a tag check, i.e., a Python \code{isinstance} test.
The sole exception is optional types, which require a tag check and a test for the none value.
Even checked dictionaries rely on tag checks.
A checked dict type with keys $\spteval_0$ and values $\spteval_1$ accepts only
checked dict values that were initialized with exactly the same key and value type.
As noted in~\cref{s:type-dynamic}, this exact-match rule applies even when $\spteval_0$
or $\spteval_1$ is the dynamic type.

@@

We end here, with materialization, rather than present a semantics
for the formalization.
After all, the main benefit of a full formal semantics is to validate
the behavior of complex expressions---and this job is
better left to the Redex mechanization, which at any rate supports a much
larger subset of Python~(\cref{s:surface}).

@@

For interested readers, we offer the following direct argument as well.
Boundaries to less-typed code would be \emph{unsound} if an untyped
(or partially-untyped) value could enter a typed context without
a validating check.
The question is thus whether all boundaries are properly guarded.
Our answer has two parts.
First, observe that every evaluation type comes with a decidable
cast~(\cref{t:cast}).
For basic types such as strings, a tag check clearly suffices.
For parameterized types such as $\sptchkdict{\spteval_0}{\spteval_1}$,
the exact-match semantics is also decidable (and also implemented with
tag checks).
Second, observe that boundaries can only relate the dynamic type to an
evaluation type; there are no significant boundaries that relate a partially
static type to an evaluation type.
This is due to the exact-match semantics for checked types and the fact that
$\sptoptional{\sptdyn}$ normalizes to the dynamic type.
Therefore casts at static occurrences of the materialization rule are
enough to protect all boundaries.

@@

The model intentionally does not cover all of Python.
Some aspects of \SP{} are left out because they are straightforward to
handle.
These include primitive types~(\cref{s:c-types}), module-level variables,
and the boundary to untyped Python~(\cref{s:py-boundary}).
%% Aspects of \SP{} that do not pertain to soundness, such as the quality of error messages,
%% are also omitted.
Other Python features are left out because their \SP{} semantics is identical
to standard Python~(\cref{s:py-boundary}).
Our model also does not cover the optimizer~(\cref{s:optimize}) because
a formal account of type-directed optimization would require substantial additional
components; namely, a model of the \SP{} bytecode and a faithful rendering of its
transformations.

@@

\SP{} can generate efficient code because it targets
the Cinder runtime, which has bytecode instructions to check types at run-time, to
construct \SP{} data structures, and to perform type-directed actions.
\Cref{t:bytecode} lists a few representative bytecode instructions.
The two instructions that express run-time checks are \bcinst{CAST}
and \bcinst{CHECK\_ARGS}.
The former checks a value against a type.
The latter is for functions and methods; it checks all inputs to a function
against their declared types.
The instructions \bcinst{BUILD\_CHECKED\_MAP} and \bcinst{TP\_ALLOC}
allocate \SP{}-specific data structures.
The first creates a checked dictionary; the second allocates an uninitialized instance
of an object~(\cref{s:vtable}).
Cinder comes with a similar instruction to build checked lists.
The remaining instructions are for optimization.
Both \bcinst{INVOKE\_FUNCTION} and \bcinst{INVOKE\_METHOD} are alternatives
to Python's dynamic call dispatch.
The former uses the classloader to quickly find the address of a function;
the latter uses a vtable lookup to find a method.
The load and store instructions for fields improve upon Python's generic attribute lookup.
In the JIT, these instructions can be further optimized to a single assembly instruction.
Lastly, the \bcinst{REFINE\_TYPE} instruction tells the JIT about the type of a local value
when it is not clear from the context.

@@

Developers have converted over 500 modules to \SP{} thus far.
Despite some initial regressions, none of the converted modules
ran slower after small rewrites~(\cref{s:code-changes})
and/or enabling the JIT on certain functions.
Only 9 modules use concrete types and only 12 modules use primitive types,
but these features delivered critical performance in key modules.

@@

\SP{} first entered the web server codebase as a replacement for a
few critical Cython modules.
Cython improved performance by compiling these modules to C,
but its partial adoption had led to an awkward workflow with a
few compiled modules spread across an interpreted codebase.
Replacing these modules with \SP{} lets programmers return to a conventional
Python workflow.
Furthermore, \SP{} types combined with Cinder resulted in a
0.7\% improvement in web server CPU efficiency.
Primitive types~(\cref{s:c-types}) were essential for matching Cython.

Later \SP{} migrations have been directed by profiling to find frequently-executed code.
During the first half of 2021, the \SP{} team identified hot
modules and proposed types plus small code changes to the maintainers of these modules~(\cref{s:code-changes}).
The accepted proposals resulted in a 1\% improvement in CPU efficiency.
During the second half of 2021, the \SP{} team applied the same
process at a larger scale.
They also modified a code-generating module to output \SP{} code;
this one change brought over four hundred generated modules into the typed fold.
Overall, the second-half changes added 2\% to the number of requests
that servers could handle per second at maximum CPU load.

@@

  \item
    \ipara{Refine some Pyre annotations}
    Well-typed Pyre code is not always well-typed \SP{} code.
    \Cref{a:pyre-limits} lists the main friction points.

@@

\Cref{t:microbenchmark} compares \SP{} to Python on public microbenchmarks.
These microbenchmarks are admittedly small and quite different than typical application code,
but they provide an auditable and reproducible way to measure performance.

@@

The 25 GitHub issues that arose from our work may be found at the following three links:

\begin{itemize}
  \item \shorturl{https://}{github.com/facebookincubator/cinder/issues/created\_by/LuKC1024} (N=20)
  \item \shorturl{https://}{github.com/facebookincubator/cinder/issues/created\_by/bennn} (N=4)
  \item \shorturl{https://}{github.com/facebookincubator/cinder/issues/64} (N=1)
  %% extra, can't remember if we inspired Dino or not: https://github.com/facebookincubator/cinder/issues/32
\end{itemize}

>>>>

%% TODO 2022-01-25 recount tests
\newcommand{\numbenchmark}{4}
\newcommand{\numSPtest}{265}
\newcommand{\numSPtotal}{802} %% SP commit 6d61575 from Jan 9
\newcommand{\numSPdiff}{537} %% (\numSPtotal - \numSPtest)
\newcommand{\CPUchange}{3.7\%}

@@

This paper reports an exception to the rule among industry-made gradual type systems.
The \SP{} team at Instagram has developed a sound type system for a subset of
Python along with a runtime system that uses soundness to drive optimizations.
The language is staged to let programmers choose between easy migrations and
full-strength optimizations, a design that we call \emph{gradual soundness}.
To a first approximation, there are three main stages:
\begin{enumerate}
  \item
    A full language of \emph{shallow types\/} that describe idiomatic
    Python code at a coarse granularity.
    Adapting an untyped module to use shallow types requires at most a few local code
    changes.
  \item
    A second level of \emph{concrete types\/} describes generic data structures
    that check the types of their elements.
    If programmers modify their code to build concrete structures instead of Python
    ones (a potentially nonlocal change), then \SP{} can perform deeper optimizations.
  \item
    A third set of progressive types enable further optimizations.
    In particular, \emph{primitive types\/} allow for unboxed arithmetic.
\end{enumerate}

@@

  \item
    \emph{Mechanization}.
    To validate the soundness of \SP{}, we mechanized a core language in PLT Redex
    and ran both property-based soundness tests and over 250 tests adapted from the \SP{} regression suite.
    The modeling effort revealed 20 significant issues in \SP{},
    four of which were soundness bugs.

@@

This paper begins with an informal description of \SP{} in two parts.
First, we present a user-oriented summary of the language~(\cref{s:tour}).
Second, we present the key ingredients of the runtime system, Cinder,
that supports \SP{}~(\cref{s:runtime}).
\Cref{s:model} uses a small formal model to introduce our Redex mechanization
and to demonstrate that \SP{} is based on a sound core.
\Cref{s:impl} notes important aspects of \SP{} that are not covered by the model.
\Cref{s:eval} evaluates \SP{}; it reports our experience with the language in
production and on public microbenchmarks.
The paper concludes with related work~(\cref{s:related})
and a final summary~(\cref{s:conclusion}).

@@

\SP{} is part of a large codebase that includes a type system, a tailored
bytecode, and a method-based JIT compiler.
In essence, \SP{} is the type system for an entirely new language.
The interface that it offers to programmers, however, replicates the
standard Python experience.
\SP{} runs Python 3.8 programs with minimal changes to their
behavior~(\cref{s:not-python}),
and it compiles code on-the-fly to support existing
IDEs and developer tools.
The advanced features of the \SP{} type system are offered on an opt-in basis
(\emph{gradual soundness})
and arranged so that programmers can begin adding types one module at a time.

@@

The \SP{} type system is a unique synthesis of ideas from the gradual typing literature
and prior work on types for Python.
Its syntax for types is based on the definitions in \PEP{}~\cite{pep484}.
Its static semantics for types is inspired by optional type checkers; in
particular, Pyre~\cite{pyre} and mypy~\cite{mypy}.
And its strategy for run-time checks is adapted from Nom~\cite{mt-oopsla-2017,mt-oopsla-2021},
a research language with compelling performance results.
\SP{}'s novelty comes from the following engineering constraints:
\begin{itemize}
  \item
    Performance is the bottom line.
    At the end of the day, \SP{} needs to make code run faster in production.
  \item
    Critical code often relies on first-order functions and objects,
    at least in the Instagram web server.
    Precise types for higher-order functions, for first-class classes, and even for
    locally-defined functions and classes are thus a low priority; instead, the
    dynamic type serves as a coarse approximation.
  \item
    The codebase in which \SP{} is being applied makes heavy use of
    \PEP{} type annotations and is regularly analyzed by Pyre.
    \SP{} is therefore compatible with the \PEP{} syntax to reduce the
    adoption burden.
    %% depth-first approach? (the alternatives are not very clear ... depth vs breadth?)
  \item
    A module-level granularity is acceptable.
    Once \SP{} is enabled for a module, it compiles all code in that module
    including expressions that have the dynamic type~(\cref{s:type-dynamic}).
    If this behavior is problematic for certain expressions, they must move
    to an untyped Python module.
\end{itemize}
%
In general, the types that \SP{} implements all enable significant optimizations
and can be validated quickly with casts.
Types that do not meet these criteria are deferred to the mature optional
type checkers that already exist for Python.
In fact, \SP{} has no short-term plans to support all \PEP{} types.

@@

\end{enumerate}
%
With this context in mind, \SP{} keeps track of whether each class is typed or untyped.
Typed classes can assume that if a method has a precise (non-dynamic)
type, then all typed overrides are subtypes of this type.
This property enables optimized dispatch from typed callers to typed methods
and optimized field reads.
Other method and field accesses typically incur one run-time check.
An extra step arises when an untyped method overrides a typed method.
In this case, \SP{} creates a wrapper around the overriding method to check
that it computes type-correct results.
% Because these wrappers perform a simple first-order check, they
% are supported by an efficiently implementation that cooperates with
% \SP{}
The wrappers are handled efficiently by a tailored implementation of vtables~(\cref{s:vtable}).

@@

For performance-critical applications, \SP{} includes a set of primitive types
that describe booleans and sized numbers.
Example types include \code{int64}, \code{uint64}, \code{double}.
There are also two special datatypes, \code{Array} and \code{Vector}, that store primitives.
The purpose of these types is to allow unboxed values and arithmetic at
runtime.
Consequently, their static semantics is a \emph{progressive}~\cite{pqk-onward-2012}
refinement over the semantics of basic Python types:
they permit fewer behaviors statically to enable the unboxing.
Operations such as \code{and} (boolean conjunction), for example, do not accept
a mix of primitive and non-primitive arguments.
These primitive types are another central aspect of gradual soundness.

@@

Relative to Python, \SP{} makes a few behavioral changes to enable
sound types and strong optimizations:
module-level variables cannot be mutated;
class attributes must be set in an \code{\_\_init\_\_} method to
ensure predictable layouts (all \PEP{} checkers agree on this);
typed attributes are resolved statically, and thus cannot be overridden with
descriptors or properties;
typed fields 
and
multiple inheritance of either typed classes or untyped children
of typed classes is forbidden.

@@

All CPython 3.8 instructions have the same behavior in Cinder.
Cinder adds instructions to help with one of three tasks:
expressing runtime checks,
initializing concrete-typed values,
or efficiently performing a standard action.
As an example of the third kind, the \code{FAST\_LEN} instruction quickly
computes the length of a built-in value, which helps to optimize loops.
\SP{} uses type information to decide where this instruction is appropriate
and applies it as an optimization.

@@

The concrete versions of built-in data structures come with both a type and an
implementation.
The implementation provides the same interface as the built-in, but uses
type tags to reject certain operations.

@@

Cinder includes a JIT (just-in-time) compiler for its bytecode.
At the moment, programmers enable the JIT by supplying a list of method names
to the compiler.
Any method can be JIT-compiled whether or not it uses \SP{} types,
though fewer types usually result in fewer optimizations.

@@

The modelling process helped uncover several critical issues in \SP{}.
This is a very important payoff given that \SP{} is running in production at
Instagram.
Overall, we made 26 bug reports~(\cref{a:github-issues}) to the \SP{} team via GitHub.
Five of these were soundness issues, one of which we could exploit to raise
a segmentation fault.
All of these soundness bugs have been fixed.
Of the remaining issues, five were relatively minor; these dealt with confusing error messages
and incorrect tests.
The remaining 16 issues report bugs in language design and implementation.
The \SP{} team has acknowledged these as bugs by applying a specific GitHub label: \lstinline{sp-correctness}.

%% critical soundness issues:
%% https://github.com/facebookincubator/cinder/issues/62
%% https://github.com/facebookincubator/cinder/issues/55
%% https://github.com/facebookincubator/cinder/issues/53
%% https://github.com/facebookincubator/cinder/issues/39
%% https://github.com/facebookincubator/cinder/issues/36

@@

The rest of this section illustrates our Redex model using a tiny
formalization to highlight the boundaries between typed and dynamic-typed
code.
\Cref{s:surface} presents a surface syntax.
\Cref{s:eval-types} explains how types get enforced at run-time.
\Cref{s:ts} argues that the overall approach toward boundaries is sound.

@@

\Cref{f:surface-types} presents a core syntax for programs.
A program is a sequence of statements;
a statement defines a variable, function, or class.
These definitions may only appear on the top level
and they all require type annotations.
Functions must have one positional argument.
Classes must declare one parent (either $\sptobject$ or another class), one field, and one method.
Expressions describe values and simple computations.
The basic values are the none value, integers,
booleans (which are the integers \code{0} and \code{1}), and strings.
There are two data structures: dictionaries and checked dictionaries~(\cref{s:checked-type}).
Other expressions describe function calls, dictionary reads and writes,
object creation, object field reads and writes, and method calls.

In Python, the syntax for expressions overlaps with the syntax of types.
For example, the Python name \code{None} is both an expression and a type,
and the \SP{} names \code{chkdict} and \code{CheckedDict} are synonyms.
\Cref{f:surface-types} does not follow these standards.
Instead, it keeps the two syntaxes distinct.
Expressions use lowercase names (\code{none}, \code{chkdict}) and types use capitalized
names (\code{None}, \code{CheckedDict}).

Surface types $\sptype$ include the dynamic type, types for the basic values, one type $\sptclass$ for every
user-defined class, and union types.
We choose to write the dynamic type as $\sptdyn$ for simplicity; technically,
\SP{} does not have a surface syntax for its dynamic
type and allows missing and unknown types~(\cref{s:type-dynamic}).
We assume that all unions are written in a flat and simplified form, e.g., that
$\sptunion{\sptint, \sptunion{\sptdyn, \sptclass_0}}$ would be normalized to $\sptdyn$.

@@

\SP{} evaluation types can all be enforced fully and efficiently
with casts.
\Cref{t:cast} describes the casts in detail.
Most call for a tag check, i.e., a Python \code{isinstance} test.
The sole exception is optional types, which require a tag check and a test for the none value.
Checked dictionaries also rely on tag checks;
a checked dict type with keys $\spteval_0$ and values $\spteval_1$ accepts only
checked dict values that were initialized with exactly the same key and value type.
As noted in~\cref{s:type-dynamic}, this exact-match rule applies even when $\spteval_0$
or $\spteval_1$ is the dynamic type.

@@

We end here, with materialization, rather than present a semantics
for the formalization.
After all, the main benefit of a full formal semantics is to validate
the behavior of complex expressions---and this job is
better left to the Redex mechanization, which supports a much
larger subset of Python~(\cref{s:surface}).

@@

For interested readers, we offer the following direct argument as well.
Boundaries to less-typed code would be \emph{unsound} if an untyped
(or partially-untyped) value could enter a typed context without
a validating check.
The question is thus whether all boundaries are properly guarded.
Our answer has two parts.
First, observe that every evaluation type comes with a decidable
cast~(\cref{t:cast}).
For basic types such as strings, a tag check clearly suffices.
For parameterized types such as $\sptchkdict{\spteval_0}{\spteval_1}$,
the exact-match semantics is also decidable (and also implemented with
tag checks).
Second, observe that boundaries can only relate the dynamic type to an
evaluation type; there are no significant boundaries that relate a partially
static type to an evaluation type.
This is due to the exact-match semantics for checked types and the fact that
$\sptoptional{\sptdyn}$ normalizes to the dynamic type.
Therefore, inserting casts at static occurrences of the materialization rule is
enough to protect all boundaries.

@@

The Redex model intentionally does not cover all of Python.
Some aspects of \SP{} are left out because they are straightforward to
handle.
These include primitive types~(\cref{s:c-types}), module-level variables,
and the boundary to untyped Python~(\cref{s:py-boundary}).
%% Aspects of \SP{} that do not pertain to soundness, such as the quality of error messages,
%% are also omitted.
Other Python features are left out because their \SP{} semantics is identical
to standard Python~(\cref{s:py-boundary}).
Our model also does not cover the optimizer~(\cref{s:optimize}) because
a formal account of type-directed optimization would require substantial additional
components; namely, a model of the \SP{} bytecode and a faithful rendering of its
transformations.

@@

\SP{} can generate efficient code because it targets
the Cinder runtime.
Cinder provides bytecode instructions that check types at run-time,
construct \SP{} data structures, and perform type-directed actions.
\Cref{t:bytecode} lists a few representative instructions.
The two instructions that express run-time checks are \bcinst{CAST}
and \bcinst{CHECK\_ARGS}.
The former checks a value against a type.
The latter is for functions and methods; it checks all inputs to a function
against their declared types.
The instructions \bcinst{BUILD\_CHECKED\_MAP} and \bcinst{TP\_ALLOC}
allocate \SP{}-specific data structures.
The first creates a checked dictionary; the second allocates an uninitialized instance
of an object~(\cref{s:vtable}).
Cinder comes with a similar instruction to build checked lists.
The remaining instructions are for optimization.
Both \bcinst{INVOKE\_FUNCTION} and \bcinst{INVOKE\_METHOD} are alternatives
to Python's dynamic call dispatch.
The former uses the classloader to quickly find the address of a function;
the latter uses a vtable lookup to find a method.
The load and store instructions for fields improve upon Python's generic attribute lookup.
In the JIT, these instructions can be further optimized to a single assembly instruction.
Lastly, the \bcinst{REFINE\_TYPE} instruction tells the JIT about the type of a local value
when it is not clear from the context.

@@

Developers have converted over 500 modules to \SP{} thus far.
Despite some initial regressions, none of the converted modules
ran slower after small rewrites~(\cref{s:code-changes})
and/or enabling the JIT on certain functions.
Only 9 modules use concrete types and only 12 modules use primitive types,
but these features delivered critical performance gains in key modules.

@@

\SP{} first entered the web server codebase as a replacement for a
few critical Cython modules.
Cython improved performance by compiling these modules to C,
but its partial adoption had led to an awkward workflow with a
few compiled modules spread across an interpreted codebase.
Replacing these modules with \SP{} let programmers return to a conventional
Python workflow.
Furthermore, \SP{} types combined with Cinder resulted in a
0.7\% improvement in web server CPU efficiency.
Primitive types~(\cref{s:c-types}) were essential for matching Cython.

Later \SP{} migrations have been directed by profiling to find frequently-executed code.
During the first half of 2021, the \SP{} team identified hot
modules and proposed types plus small code changes to the maintainers of these modules~(\cref{s:code-changes}).
The accepted proposals resulted in a 1\% improvement in CPU efficiency.
During the second half of 2021, the \SP{} team applied the same
process at a larger scale.
They also modified a code-generating module to output \SP{} code;
this one change resulted in over four hundred generated typed modules.
Overall, the second-half changes added 2\% to the number of requests
that servers could handle per second at maximum CPU load.

@@

  \item
    \ipara{Refine some Pyre annotations}
    Well-typed Pyre code is not always well-typed \SP{} code.
    \Cref{a:pyre-limits} lists the main points of friction.

@@

\Cref{t:microbenchmark} compares \SP{} to Python on public microbenchmarks~\cite{spbench}.
These microbenchmarks are admittedly small and quite different than typical application code,
but they provide an auditable and reproducible way to measure performance.

@@

The 26 GitHub issues that arose from our work may be found at the following three links.
A list of issue numbers follows each link.

\begin{itemize}
  \item \shorturl{https://}{github.com/facebookincubator/cinder/issues/created\_by/LuKC1024} (N=20)
    \subitem \code{\#37, 39, 40, 41, 42, 43, 45, 46, 47, 49, 50, 51, 52, 53, 59, 60, 61, 62, 63, 65}
  \item \shorturl{https://}{github.com/facebookincubator/cinder/issues/created\_by/bennn} (N=5)
    \subitem \code{\#35, 36, 54, 55, 71}
  \item \shorturl{https://}{github.com/facebookincubator/cinder/issues/64} (N=1)
    \subitem \code{\#64}
  %% extra, can't remember if we inspired Dino or not: https://github.com/facebookincubator/cinder/issues/32
\end{itemize}

\noindent{
  \newcommand{\spsoundness}[1]{\href{https://github.com/facebookincubator/cinder/issues/#1}{\code{#1}}}
The critical soundness issues were:
\code{\#}\spsoundness{36},
         \spsoundness{39},
         \spsoundness{53},
         \spsoundness{55}, and
         \spsoundness{62}.
}
